### 2015-02-21 paper/paper.tex
calling from python

\documentclass[proceed]{article}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{multicol} \usepackage{fancyheadings} \usepackage{pdfpages}
\setlength{\emergencystretch}{10em}

\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}

\newenvironment{proof}[1][Proof]{\begin{trivlist}
\item[\hskip \labelsep {\bfseries #1}]}{\end{trivlist}}
\newenvironment{definition}[1][Definition]{\begin{trivlist}
\item[\hskip \labelsep {\bfseries #1}]}{\end{trivlist}}
\newenvironment{example}[1][Example]{\begin{trivlist}
\item[\hskip \labelsep {\bfseries #1}]}{\end{trivlist}}
\newenvironment{remark}[1][Remark]{\begin{trivlist}
\item[\hskip \labelsep {\bfseries #1}]}{\end{trivlist}}

\DeclareMathOperator*{\Exp}{\mathbb{E}}
\DeclareMathOperator*{\Prob}{\mathbf{P}}

\title{Qualitative Probabilistic Programming}
\author {}
\begin{document}

  \maketitle

  \begin{abstract}
    Probabilistic programs with unknown functions
  \end{abstract}

  \section{Introduction}

  Probabilistic programming languages have made it easier to specify Bayesian
  models as programs.  Once a probabilistic model is written as a probabilistic
  program, it is possible to use generic inference algorithms (such as
  Metropolis Hastings) to perform inference.
    - Church example

  In this paper, we discuss a feature of probabilistic programming languages
  that makes it easy to specify models containing unknown functions.
  Specifically, we allow programs to specify that some random functions
  are unknown.  The system will fill in these unknown functions with
  some reasonable default class of distributions based on the types and
  proceed to infer the parameters to this distribution using the
  expectation maximization algorithm.

  This feature has multiple advantages.  First, it is easier to write
  models without knowing about the class of models being used.  This should
  make probabliistic programming more accessible to non-experts.  Secondly,
  parameter inference is more efficient if specialized algorithms are used
  rather than the generic algorithms used to infer other random variables
  (such as Metropolis Hastings).

  We define an example probabilistic programming language with this feature
  (Quipp) and show how it can be used to write machine learning models
  concisely.

  - Motivation for Quipp
    - Explanation of "unknown functions"
    - Writing machine learning algorithms as probabilistic programs
    - Accessibility to non-experts
    - Comparison to existing probabilistic programming languages
      - In other languages, use random variables for parameters
      - Random variables slower because they are updated independently

  \section{Syntax}

  \section{Algorithm}

    To perform inference of both latent variables and parameters, we first
    translate the program to a factor graph.  Next, we run the
    expectation maximization algorithm on this graph, iterating stages of
    estimating latent variables using Metropolis Hastings and parameter
    inference using Newton's method.

  - EM algorithm
    - E step is Metropolis Hastings as in probabilistic programming
    - M step is parameter optimization
  - Types
    - Brief overview of exponential families
    - Gaussian
    - Categorical
    - Other exponential families could be added easily
    - Some algebraic data types
      - Tuples
      - Either
  - Conversion to factor graph
    - Different types of nodes (exponential family)
    - Different types of factors
      - Constant factors
      - Unifying values (e.g. conditioning on compound value equaling something)
      - Factors from known random functions (e.g. gaussian)
      - Factors from unknown function calls
    - Example: clustering conversion to factor graph
  - Inference in factor graph
    - Metropolis hastings
    - Proposal distribution inferred from nearby nodes

  \section{Examples}

  \section{Discussion}

  - Comments about results
  - Future directions
    - Unbounded models
    - Recursive algebraic data types

\end{document}

### 2015-02-21 src/Main_old.hs

@@ -1,0 +1,66 @@
+module Main where
+
+import Data.Time.Clock (getCurrentTime, utctDayTime)
+import Control.Monad (forM_)
+import Debug.Trace
+import Control.Monad (liftM)
+import Control.Monad.Trans (lift)
+import Data.Maybe (fromJust)
+import Data.Random (RVar, RVarT, runRVarTWith, StdRandom(StdRandom))
+import Text.ParserCombinators.Parsec
+import Data.Function (on)
+import Data.Map (Map)
+import qualified Data.Map as Map
+import Data.String.Utils (split)
+
+import Quipp.ExpFam
+import Quipp.Factor
+import Quipp.Vmp
+import Quipp.Util
+import Quipp.Value
+import Quipp.GraphBuilder
+import Quipp.Parser
+import Quipp.Interpreter
+import Quipp.ParamInference
+
+fPairGraphValue x y = FCompoundGraphValue "Pair" [x, y]
+
+example1dClustering = ("1d_clustering.quipp", map (FValueGraphValue . DoubleValue) [
+  1.0, 1.1, 1.2, 1.4, 6.7, 7.9, 8.9, 5.0
+  ])
+
+example2dClustering points = ("nd_clustering.quipp", map (uncurry (fPairGraphValue `on` (FValueGraphValue . DoubleValue))) points)
+--  (1.5, 11.0), (2.2, 10.5), (1.4, 9.8), (2.5, 20.5), (5.3, 22.7),
+--  (1.1, 18.6), (2.3, 21.7), (3.5, 23.8), (4.3, 25.3), (5.5, 28.0)
+--  ])
+
+irisData = do
+  strs <- readFile "data/iris.data.txt"
+  return [(read a :: Double, read c :: Double) | line <- lines strs, line /= "", let [a, b, c, d, e] = split "," line]
+
+main = do
+  -- let (filename, samples) = example2dClustering
+  datapoints <- irisData
+  let (filename, samples) = example2dClustering datapoints
+  -- print datapoints
+  prelude <- readFile "Quipp/prelude.quipp"
+  contents <- readFile $ "examples/" ++ filename
+  let fullSource = prelude ++ contents
+  let resultExpr =
+        case parse toplevel "FILE" fullSource of
+          Left err -> error $ show err
+          Right result -> result
+      builder = interpretExpr defaultContext Map.empty resultExpr
+      -- (template, result) = runGraphBuilder builder
[…]

### 2015-02-21 src/python/callhaskell.py

@@ -1,8 +1,46 @@
-
+import json
+import uuid
+import subprocess
 
 
 """
 Utilities for calling the Haskell code.
+
+JSON:
+
+expfam:
+  'gaussian', 'bernoulli'
+
+template:
+    {vars: [expfam],
+     randFuns: [
+       {argExpFams: [expfam],
+        resExpFam: expfam
+       }
+     ],
+     factors: [
+       {factor:
+        argVarIds: [int],
+        {type: 'randFun', randFunId: int}
+        ||
+        {type: 'constant', value: *}}
+     ]
+    }
+
+    This only has to be encoded by Python and decoded by Haskell!
+
+params:
+  {<int>: {
+    base: [double],
+    matrix: [[double]]
+  }
+
+  Python does not need to know anything about this.
+
+state:
+  {<int>: *}
+
+  Both Python and Haskell must know about this.
 
 pyRandTemplateParams(templ)
   generate random parameters for a given template
@@ -17,6 +55,53 @@ pyStepEM(templ, state, params)
   return [newState, newParams]
 """
 
+def temp_file_name(prefix=''):
+    return '/tmp/' + prefix + '_' + str(uuid.uuid1())
 
+def run_hs(*args):
+    subprocess.call(['../../dist/build/quipp_main/quipp_main'] + list(args))
 
+def dump_template(templ):
+    templ_file = temp_file_name('template')
+    json.dump(templ, open(templ_file, 'w'))
+    return templ_file
 
+def dump_state(state):
+    state_file = temp_file_name('state')
+    json.dump(state, open(state_file, 'w'))
+    return state_file
+
+def dump_params(params):
+    params_file = temp_file_name('params')
+    json.dump(params, open(params_file, 'w'))
+    return params_file
+
+def load_contents(fname):
+    return json.load(open(fname))
+
+def hs_rand_template_params(templ):
+    params_file = temp_file_name('params')
+    run_hs('randTemplateParams', dump_template(templ), params_file)
+    return params_file
+
+def hs_sample_bayes_net(templ, params):
+    state_file = temp_file_name('state')
+    run_hs('randTemplateParams', dump_template(templ), dump_params(params), state_file)
+    return state_file
+
+def hs_init_em(templ):
+    state_file = temp_file_name('state')
+    params_file = temp_file_name('params')
+    run_hs('initEM', dump_template(templ), state_file, params_file)
+    return (load_contents(state_file), load_contents(params_file))
[…]

### 2015-02-21 src/python/examples/lda.py

@@ -1,0 +1,17 @@
+from quipp import *
+
+n_classes = 10
+n_words = 10000
+n_words_per_document = 1000
+
+ClassType = Categorical(n_classes)
+WordType = Categorical(n_words)
+
+class_to_word = rand_function(ClassType, WordType)
+
+def sample():
+    which_class = uniform_categorical(n_classes)
+    words = [class_to_word(which_class) for i in range(n_words_per_document]
+    return (which_class, words)
+
+run_example(sample)

### 2015-02-21 src/python/graphbuilder.py

@@ -19,103 +19,101 @@ class RandFunId(object):
     assert isinstance(id, int)
     self.state = state
     self.id = id
-    
+
 class GraphState(object):
 
   def __init__(self):
     self.vars = []
-    self.randFuns = []
+    self.rand_funs = []
     self.factors = []
-    self.varReplacements = {}
+    self.var_replacements = {}
 
-  def newVar(self, expfam):
+  def new_var(self, expfam):
     varid = len(self.vars)
     self.vars.append(expfam)
     return VarId(self, varid)
 
-  def resolveVar(self, varid):
-    if varid.id in self.varReplacements:
-      return self.varReplacements[varid.id]
+  def resolve_var(self, varid):
+    if varid.id in self.var_replacements:
+      return self.var_replacements[varid.id]
     else:
       return varid
 
-  def unifyVars(self, a, b):
+  def unify_vars(self, a, b):
     def replace(v):
       return a if v == b else v
 
-    def replaceInFactor(fac):
-      (facInfo, args) = fac
-      return (facInfo, map(replace, args))
+    def replace_in_factor(fac):
+      (fac_info, args) = fac
+      return (fac_info, map(replace, args))
 
-    self.varReplacements[a.id] = b
-    self.factors = map(replaceInFactor, self.factors)
+    self.var_replacements[a.id] = b
+    self.factors = map(replace_in_factor, self.factors)
     return a
 
-  def unifyValues(self, typ, a, b):
+  def unify_values(self, typ, a, b):
     # TODO: this fails for e.g. Maybe
-    avars = typ.embeddedVars(a)
-    bvars = typ.embeddedVars(b)
+    avars = typ.embedded_vars(a)
+    bvars = typ.embedded_vars(b)
     assert len(avars) == len(bvars)
     for av, bv in zip(avars, bvars):
-      self.unifyVars(av, bv)
+      self.unify_vars(av, bv)
     return a
 
-  def newFactor(self, facInfo, args):
+  def new_factor(self, fac_info, args):
     facid = len(self.factors)
-    self.factors.append((facInfo, map(self.resolveVar, args)))
+    self.factors.append((fac_info, map(self.resolve_var, args)))
     return FactorId(self, facid)
 
-  def newRandFun(self, argExpFams, resExpFam):
-    rfid = len(self.randFuns)
[…]

### 2015-02-24 src/Main.hs
python examples

@@ -1,66 +1,147 @@
+{-# LANGUAGE FlexibleInstances #-}
 module Main where
 
-import Data.Time.Clock (getCurrentTime, utctDayTime)
-import Control.Monad (forM_)
-import Debug.Trace
 import Control.Monad (liftM)
-import Control.Monad.Trans (lift)
-import Data.Maybe (fromJust)
-import Data.Random (RVar, RVarT, runRVarTWith, StdRandom(StdRandom))
-import Text.ParserCombinators.Parsec
-import Data.Function (on)
-import Data.Map (Map)
+import Control.Applicative ((<$>), (<*>))
+import Data.Map (Map, (!))
 import qualified Data.Map as Map
-import Data.String.Utils (split)
+import System.Environment (getArgs)
+import qualified Text.JSON as JSON
 
+import Quipp.BayesNet
 import Quipp.ExpFam
 import Quipp.Factor
-import Quipp.Vmp
+import Quipp.GraphBuilder
 import Quipp.Util
 import Quipp.Value
-import Quipp.GraphBuilder
-import Quipp.Parser
-import Quipp.Interpreter
-import Quipp.ParamInference
+import Quipp.Vmp
 
-fPairGraphValue x y = FCompoundGraphValue "Pair" [x, y]
 
-example1dClustering = ("1d_clustering.quipp", map (FValueGraphValue . DoubleValue) [
-  1.0, 1.1, 1.2, 1.4, 6.7, 7.9, 8.9, 5.0
-  ])
 
-example2dClustering points = ("nd_clustering.quipp", map (uncurry (fPairGraphValue `on` (FValueGraphValue . DoubleValue))) points)
---  (1.5, 11.0), (2.2, 10.5), (1.4, 9.8), (2.5, 20.5), (5.3, 22.7),
---  (1.1, 18.6), (2.3, 21.7), (3.5, 23.8), (4.3, 25.3), (5.5, 28.0)
---  ])
+zip1_2 :: [a] -> [(b, c)] -> [(a, b, c)]
+zip1_2 xs yzs = zipWith (\x (y, z) -> (x, y, z)) xs yzs
 
-irisData = do
-  strs <- readFile "data/iris.data.txt"
-  return [(read a :: Double, read c :: Double) | line <- lines strs, line /= "", let [a, b, c, d, e] = split "," line]
+decodeExpFam :: String -> ExpFam Value
+decodeExpFam "gaussian" = gaussianValueExpFam
[…]

### 2015-02-24 src/Quipp/ExpFam.hs

@@ -194,6 +194,7 @@ timesLikelihood (NatParam n1) (NatParam n2)
 
 productLikelihoods :: Eq v => [Likelihood v] -> Maybe (Likelihood v)
 productLikelihoods (l:ls) = foldlM timesLikelihood l ls
+productLikelihoods [] = error "Cannot take the product of no likelihoods"
 
 sampleLikelihood :: ExpFam v -> Likelihood v -> RVar v
 sampleLikelihood _ (KnownValue v) = return v

### 2015-02-24 src/Quipp/Factor.hs

@@ -151,7 +151,7 @@ data FactorGraphTemplate v = FactorGraphTemplate {
 } deriving Show
 
 makeFactorGraphTemplate :: [(VarId, ExpFam v)] -> [(RandFunId, ExpFam v, [ExpFam v])] -> [(FactorId, Either (Factor v) RandFunId, [VarId])] -> FactorGraphTemplate v
-makeFactorGraphTemplate vars randfuns factors = FactorGraphTemplate (Map.fromList fgv) (Map.fromList fgrf) (Map.fromList fgf)
+makeFactorGraphTemplate vars randfuns factors = traceShow (vars, fgv) $ FactorGraphTemplate (Map.fromList fgv) (Map.fromList fgrf) (Map.fromList fgf)
   where fgv = [(varid, (ef, [factorid | (factorid, _, vars) <- factors, elem varid vars])) | (varid, ef) <- vars]
         fgrf = [(rfid, (ef, featureEfs, [factorid | (factorid, Right rfid', _) <- factors, rfid' == rfid])) | (rfid, ef, featureEfs) <- randfuns]
         fgf = [(factorid, (fac, vars)) | (factorid, fac, vars) <- factors]

### 2015-02-24 src/Quipp/PyExports.hs

@@ -1,63 +1,0 @@
-{-# LANGUAGE FroeignFunctionInterface, TemplateHaskell #-}
-module PyExports where
-
-import System.IO.Unsafe (unsafePerformIO)
-import Text.JSON.Generic (encodeJSON, decodeJSON)
-
-import Quipp.BayesNet
-import Quipp.ExpFam
-import Quipp.Factor
-import Quipp.GraphBuilder
-import Quipp.Interpreter
-import Quipp.Util
-import Quipp.Value
-import Quipp.Vmp
-
-
-decodeTemplate :: String -> FactorGraphTemplate Value
-
-decodeParams :: String -> FactorGraphParams
-decodeParams = decodeJSON
-
-encodeParams :: FactorGraphParams -> String
-encodeParams = encodeJSON
-
-decodeState :: String -> FactorGraphState Value
-decodeState = decodeJSON
-
-encodeState :: FactorGraphState Value -> String
-encodeState = encodeJSON
-
-pyRandTemplateParams :: String -> IO String
-
-pySampleBayesNet :: String -> String -> IO String
-
-pyInitEM :: String -> IO [String]
-
-pyStepEM :: String -> String -> String -> IO String
-
-initHaPy
-pythonExport 'pyRandTemplateParams
-pythonExport 'pySampleBayesNet
-pythonExport 'pyInitEM
-pythonExport 'pyStepEM
-
-
-"""
-Utilities for calling the Haskell code.
-
-pyRandTemplateParams(templ)
-  generate random parameters for a given template
-
-pySampleBayesNet(templ, params)
-  given template and parameters, return a sample map of varids to value
-
-pyInitEM(templ)
-  generate [state, params]
-
-pyStepEM(templ, state, params)
-  return [newState, newParams]
-"""
-
-
-

### 2015-02-24 src/python/callhaskell.py

@@ -55,6 +55,18 @@ pyStepEM(templ, state, params)
   return [newState, newParams]
 """
 
+def state_to_varvals(state):
+  m = {}
+  for (v, s) in state:
+    m[v] = s[u'Left']
+  return m
+
+def varvals_to_state(varvals):
+  s = []
+  for k in varvals:
+    s.append([k, {u'Left': varvals[k]}])
+
+
 def temp_file_name(prefix=''):
     return '/tmp/' + prefix + '_' + str(uuid.uuid1())
 
@@ -77,17 +89,17 @@ def dump_params(params):
     return params_file
 
 def load_contents(fname):
-    return json.load(open(fname))
+  return json.load(open(fname))
 
 def hs_rand_template_params(templ):
     params_file = temp_file_name('params')
     run_hs('randTemplateParams', dump_template(templ), params_file)
-    return params_file
+    return load_contents(params_file)
 
 def hs_sample_bayes_net(templ, params):
     state_file = temp_file_name('state')
-    run_hs('randTemplateParams', dump_template(templ), dump_params(params), state_file)
-    return state_file
+    run_hs('sampleBayesNet', dump_template(templ), dump_params(params), state_file)
+    return load_contents(state_file)
 
 def hs_init_em(templ):
     state_file = temp_file_name('state')
@@ -101,7 +113,7 @@ def hs_step_em(templ, state, params):
     run_hs('stepEM', dump_template(templ), state_file, params_file)
     return (load_contents(state_file), load_contents(params_file))
 
-__all__ = ['hs_rand_template_params', 'hs_sample_bayes_net', 'hs_init_em', 'hs_step_em']
+__all__ = ['hs_rand_template_params', 'hs_sample_bayes_net', 'hs_init_em', 'hs_step_em', 'state_to_varvals', 'varvals_to_state']
 
 if __name__ == '__main__':
     print hs_init_em({'vars': [], 'randFuns': [], 'factors': []})

### 2015-02-24 src/python/examples/nd_clustering.py

@@ -5,10 +5,13 @@ nclusters = 3
 PointType = Vector(dim, Double)
 ClusterType = Categorical(nclusters)
 
+get_cluster = rand_function(ClusterType)
+
 get_point = rand_function(ClusterType, PointType)
 
 def sample():
-  cluster = Uniform(nclusters)
+  # cluster = Uniform(nclusters)
+  cluster = get_cluster()
   return (cluster, get_point(cluster))
 
 run_example(sample)

### 2015-02-24 src/python/graphbuilder.py

@@ -1,4 +1,15 @@
-import Queue
+from callhaskell import *
+
+class Queue(object):
+
+  def __init__(self, items=None):
+    if items is None:
+      self.items = []
+    else:
+      self.items = list(reversed(items))
+
+  def dequeue(self):
+    return self.items.pop()
 
 
 class VarId(object):
@@ -23,15 +34,17 @@ class RandFunId(object):
 class GraphState(object):
 
   def __init__(self):
-    self.vars = []
+    self.vars = {}
+    self.var_count = 0
     self.rand_funs = []
     self.factors = []
     self.var_replacements = {}
 
   def new_var(self, expfam):
-    varid = len(self.vars)
-    self.vars.append(expfam)
-    return VarId(self, varid)
+    varid = self.var_count
+    self.var_count += 1
+    self.vars[varid] = expfam
+    return VarId(self, varid, expfam)
 
   def resolve_var(self, varid):
     if varid.id in self.var_replacements:
@@ -47,7 +60,8 @@ class GraphState(object):
       (fac_info, args) = fac
       return (fac_info, map(replace, args))
 
-    self.var_replacements[a.id] = b
+    self.var_replacements[b.id] = a
+    del self.vars[b.id]
     self.factors = map(replace_in_factor, self.factors)
     return a
 
@@ -79,8 +93,8 @@ class GraphState(object):
 
   def new_const_factor(self, varid, value):
     varid = self.resolve_var(varid)
-    ef = self.vars[varid]
-    return new_factor({'type': 'constant', 'value': value})
+    ef = self.vars[varid.id]
+    return self.new_factor({'type': 'constant', 'expFam': ef, 'value': value}, [varid])
 
   def new_const_var(self, ef, value):
     varid = self.new_var(ef)
@@ -88,20 +102,20 @@ class GraphState(object):
     return varid
 
   def rand_function(self, arg_types, res_type):
-    arg_tuple_type = Tuple(arg_types)
+    arg_tuple_type = Tuple(*arg_types)
     rfids = [self.new_rand_fun(arg_tuple_type.exp_fams(), ef) for ef in res_type.exp_fams()]
     def rf(*args):
       assert len(args) == len(arg_types)
       arg_vars = arg_tuple_type.embedded_vars(tuple(args))
       res_vars = [self.new_sample_from_rand_fun(rfid, arg_vars) for rfid in rfids]
-      return res_type.vars_to_value(res_vars)
[…]

### 2015-02-25 paper/paper.tex
kay

@@
   \begin{abstract}
-    Probabilistic programs with unknown functions
+    In probabilistic programs, sometimes it is difficult to specify the correct
+    parameterized family of distributions.  We explore an extension to
+    probabilistic programming languages that allows programmers to mark some
+    distributions as unspecified.  Then, we can fill in the distribution with
+    some family and infer parameters.
   \end{abstract}
@@
   Probabilistic programming languages have made it easier to specify Bayesian
-  models as programs.  Once a probabilistic model is written as a probabilistic
+  models as programs.  Once a model is written as a probabilistic
   program, it is possible to use generic inference algorithms (such as
   Metropolis Hastings) to perform inference.
-    - Church example
 
   In this paper, we discuss a feature of probabilistic programming languages
-  that makes it easy to specify models containing unknown functions.
-  Specifically, we allow programs to specify that some random functions
+  that makes it easier to specify some models.
+  Specifically, we allow the programmer to specify that some random functions
   are unknown.  The system will fill in these unknown functions with
-  some reasonable default class of distributions based on the types and
+  some reasonable default family of distributions based on the types and
   proceed to infer the parameters to this distribution using the
@@
 
+  In an ordinary probabilistic programming language (such as Church),
+  it would be possible to treat parameters as random variables.  This
+  would allow ordinary inference algorithms to infer parameters.  However,
+  there are advantages of having unknown functions as a feature
+  in the language.
+  First, it is easier to
+  write programs without knowing the details of different distributions.
+  Second, specialized algorithms can be used to make parameter inference
+  faster.
+
+  Inference in these models is performed using the expectation-maximization
+  algorithm, with alternating steps of inferring latent variables and optimizing
+  parameters.
+
   - Motivation for Quipp
@@
 
-  \section{Algorithm}
+  Here is an example of a Quipp program to perform clustering:
+
+  \begin{verbatim}
+  PointType = Vector(2, Double)
+  ClusterType = Categorical(3)
+
+  get_cluster = rand_function(ClusterType)
+  get_point = rand_function(ClusterType, PointType)
+
+  def sample():
+    cluster = get_cluster()
+    return (cluster, get_point(cluster))
+  \end{verbatim}
+
+  Notice that we declared two types (\texttt{PointType} and \texttt{ClusterType})
+  and two random functions (\texttt{get\_cluster} and \texttt{get\_point}).
+  Type annotations are necessary for random functions.  The type
+  \texttt{Vector(2, Double)} expresses the fact that the points are 2-dimensional,
+  and the type \texttt{Categorical(3)} expresses the fact that there are 3
+  possible clusters (so a cluster may be either 0, 1, or 2).  We do not know
+  the distribution over clusters or over points given a cluster.
+  We will fill in the \texttt{get\_cluster} function with
+  a learned categorical distribution, and we will fill in the
+  \texttt{get\_point} function with a learned unknown linear function with noise.
+
+  \section{Class of Distributions}
+
+    For unknown functions, the class of functions used is a kind
+    of generalized linear model.  We assume that the result
+    is distributed from some exponential family.  We label
+    some subset of the sufficient statistics as \emph{features}.  The natural
+    parameters corresponding to non-features are constant, while natural
+    parameters corresponding to features are determined as an affine
+    function of the features of the arguments.
+
+    Let $\psi_x$ be a matrix such that $\psi \phi(x)$ is a vector of the
+    features of $x$.  Then
+
+    $$p_{\mathbf{N}}(y | x) = \exp(\psi(x)^T N \phi(y) - g(N^T \psi(X)))$$
+
+    Given $(x, y)$ samples, parameter estimation to maximize log probability is a convex
+    problem:
+
+    \begin{align*}
+      \log p_{\mathbf{N}}(y | x) = \psi(x)^T N \phi(y) - g(N^T \psi(X))
+    \end{align*}
+    since the log probability function is concave when $g$ is convex (as it is
+    for any exponential family distribution).
+
+    Therefore, it is possible to use Newton's method to optimize the parameters.
+
+  \section{Inference}
 
@@
 
-  - EM algorithm
-    - E step is Metropolis Hastings as in probabilistic programming
-    - M step is parameter optimization
-  - Types
-    - Brief overview of exponential families
-    - Gaussian
-    - Categorical
-    - Other exponential families could be added easily
-    - Some algebraic data types
-      - Tuples
-      - Either
-  - Conversion to factor graph
-    - Different types of nodes (exponential family)
-    - Different types of factors
-      - Constant factors
-      - Unifying values (e.g. conditioning on compound value equaling something)
-      - Factors from known random functions (e.g. gaussian)
-      - Factors from unknown function calls
-    - Example: clustering conversion to factor graph
-  - Inference in factor graph
-    - Metropolis hastings
-    - Proposal distribution inferred from nearby nodes
+    In the factor graph, there are 2 types of variables: real numbers
+    and categorical values.
+    In Quipp, there are more value types than this.  Specifically,
+    in addition to real numbers and categorical types, tuples
+    are also supported.
+    To translate a Quipp program to a factor graph, we
+    map each value to a list of variables, and create factors
+    over all these variables.
+
+    There are a few different types of factors:
+    \begin{enumerate}
+      \item Factors representing known distributions
+      \item Factors representing unknown distribution
+      \item Factors asserting that a value is equal to some constant
+    \end{enumerate}
+
+    How is inference in the factor graph performed?  We use the Metropolis Hastings
+    algorithm, updating one node at a time.  To get a proposal for a given node,
+    we approximate the local distribution using message passing (similar to
+    in variational message passing).  Specifically, for each factor
+    $f(x_1, x_2, ..., x_n)$ and each $1 \leq i \leq n$, we send a message
+    to variable $x_i$ that approximates $x_i \mapsto f(x_1, x_2, ..., x_n)$.
 

### 2015-02-25 src/Main.hs

@@ -1,6 +1,7 @@
 {-# LANGUAGE FlexibleInstances #-}
 module Main where
 
+import Debug.Trace (trace, traceShow)
 import Control.Monad (liftM)
 import Control.Applicative ((<$>), (<*>))
 import Data.Map (Map, (!))
@@ -21,13 +22,15 @@ import Quipp.Vmp
 zip1_2 :: [a] -> [(b, c)] -> [(a, b, c)]
 zip1_2 xs yzs = zipWith (\x (y, z) -> (x, y, z)) xs yzs
 
-decodeExpFam :: String -> ExpFam Value
-decodeExpFam "gaussian" = gaussianValueExpFam
-decodeExpFam "bernoulli" = boolValueExpFam
+decodeExpFam :: JSON.JSValue -> ExpFam Value
+decodeExpFam ef = case ef !: "type" of
+  "gaussian" -> gaussianValueExpFam
+  "bernoulli" -> boolValueExpFam
+  "categorical" -> categoricalValueExpFam (ef !: "n")
 
 getResult :: JSON.Result a -> a
 getResult (JSON.Ok a) = a
-getResult (JSON.Error str) = error str
+getResult (JSON.Error str) = error $ "Error decoding: " ++ str
 
 decodeJSON :: JSON.JSON a => String -> a
 decodeJSON = getResult . JSON.decode
@@ -36,7 +39,10 @@ infixl 9 !:
 (!:) :: JSON.JSON b => JSON.JSValue -> String -> b
 (JSON.JSObject obj) !: key = case Map.lookup key (Map.fromList (JSON.fromJSObject obj)) of
   Nothing -> error ("Key not found: " ++ key ++ " in " ++ show obj)
-  Just result -> getResult $ JSON.readJSON result
+  Just result -> case JSON.readJSON result of
+    JSON.Ok a -> a
+    JSON.Error str -> error $ "Error looking up " ++ key ++ " in " ++ show obj ++ ": " ++ str
+other !: key = error ("Not object: " ++ show other)
 
 decodeTemplate :: String -> FactorGraphTemplate Value
 decodeTemplate str = makeFactorGraphTemplate
@@ -52,10 +58,12 @@ decodeTemplate str = makeFactorGraphTemplate
           let facinfo = fac !: "factor" in
           (case facinfo !: "type" of
              "randFun" -> Right (facinfo !: "id")
-             "constant" -> Left $ constFactor (decodeExpFam (facinfo !: "expFam")) $ case facinfo !: "expFam" of
[…]

### 2015-02-25 src/Quipp/ExpFam.hs

@@ -1,4 +1,4 @@
-{-# LANGUAGE RankNTypes, TypeFamilies, NoMonomorphismRestriction #-}
+{-# LANGUAGE RankNTypes, TypeFamilies, NoMonomorphismRestriction, ScopedTypeVariables #-}
 
 module Quipp.ExpFam (ExpFam(ExpFam, expFamName, expFamD, expFamSufStat, expFamG, expFamSample, expFamRandomNatParam),
                      expFamFeaturesD, expFamSufStatToFeatures, expFamFeaturesToSufStat,
@@ -144,19 +144,14 @@ groupSamplesByFeatures samps =
 -- samples with weight not equal to 1.
 expFamMLE :: ExpFam a -> [(Double, [Double], {- [[Double]], -} [Double])] -> Params Double -> [Params Double]
 expFamMLE fam samples etaStart | expFamName fam == "gaussian" =
-  let xs = [x | (_, [x], _) <- samples]
-      ys = [y | (_, _, [y, _]) <- samples]
-      xmean = mean xs
-      ymean = mean ys
-      xvar = variance xs
-      yvar = variance ys
-      covar = covariance (zip xs ys)
-      slope = covar / xvar
-      yint = ymean - slope * xmean
-      r2 = covar^2 / (xvar * yvar)
-      residVar = (1 - r2) * yvar
-  in trace ("xmean " ++ show xmean ++ " r2 " ++ show r2) $
-    repeat $ if isNaN covar then etaStart else ([yint / residVar, -1 / (2 * residVar)], [[slope / residVar]])
+  let nxs = let (_, x, _) = head samples in length x
+      xss :: Matrix Double = [1:x | (_, x, _) <- samples]
+      ys :: [Double] = [y | (_, _, [y, _]) <- samples]
+      beta :: [Double] = matInv (matMul (transpose xss) xss) `matMulByVector` matMulByVector (transpose xss) ys
+      predYs :: [Double] = map (dotProduct beta) xss
+      resid :: Double = mean [(y - p)^2 | (y, p) <- zip ys predYs]
+      resid' = max 0.0001 resid
+  in repeat ([head beta / resid', -1 / (2 * resid')], [map (/ resid') (tail beta)])
 
 expFamMLE fam samples etaStart = -- trace ("\nexpFamMLE " ++ show samples) $
   let samples' = groupSamplesByFeatures samples
[…]

### 2015-02-25 src/Quipp/Factor.hs

@@ -151,7 +151,7 @@ data FactorGraphTemplate v = FactorGraphTemplate {
 } deriving Show
 
 makeFactorGraphTemplate :: [(VarId, ExpFam v)] -> [(RandFunId, ExpFam v, [ExpFam v])] -> [(FactorId, Either (Factor v) RandFunId, [VarId])] -> FactorGraphTemplate v
-makeFactorGraphTemplate vars randfuns factors = traceShow (vars, fgv) $ FactorGraphTemplate (Map.fromList fgv) (Map.fromList fgrf) (Map.fromList fgf)
+makeFactorGraphTemplate vars randfuns factors = FactorGraphTemplate (Map.fromList fgv) (Map.fromList fgrf) (Map.fromList fgf)
   where fgv = [(varid, (ef, [factorid | (factorid, _, vars) <- factors, elem varid vars])) | (varid, ef) <- vars]
         fgrf = [(rfid, (ef, featureEfs, [factorid | (factorid, Right rfid', _) <- factors, rfid' == rfid])) | (rfid, ef, featureEfs) <- randfuns]
         fgf = [(factorid, (fac, vars)) | (factorid, fac, vars) <- factors]

### 2015-02-25 src/Quipp/Util.hs

@@ -112,6 +112,12 @@ dotProduct x y = sum (zipWith (*) x y)
 matMulByVector :: Num a => Matrix a -> [a] -> [a]
 matMulByVector m v = map (dotProduct v) m
 
+matMul :: Num a => Matrix a -> Matrix a -> Matrix a
+matMul m1 m2 = map (matMulByVector m1) (transpose m2)
+
+matInv :: Matrix Double -> Matrix Double
+matInv = Mat.toLists . pinv . Mat.fromLists
+
 linSolve :: Matrix Double -> [Double] -> [Double]
 linSolve mat d =
   matMulByVector (Mat.toLists $ pinv $ Mat.fromLists mat) d

### 2015-02-25 src/Quipp/Value.hs

@@ -3,7 +3,7 @@ module Quipp.Value where
 import Quipp.ExpFam
 
 
-data Value = DoubleValue Double | BoolValue Bool deriving (Eq, Ord, Show)
+data Value = DoubleValue Double | BoolValue Bool | CategoricalValue Int deriving (Eq, Ord, Show)
 
 fromDoubleValue (DoubleValue a) = a
 doublePromoter = (DoubleValue, fromDoubleValue)
@@ -11,4 +11,6 @@ gaussianValueExpFam = promoteExpFam doublePromoter gaussianExpFam
 
 boolPromoter = (BoolValue . (== 1), \(BoolValue x) -> if x then 1 else 0)
 boolValueExpFam = promoteExpFam boolPromoter (categoricalExpFam 2)
+
+categoricalValueExpFam n = promoteExpFam (CategoricalValue, \(CategoricalValue x) -> x) (categoricalExpFam n)
 

### 2015-02-25 src/python/graphbuilder.py

@@ -1,3 +1,4 @@
+import math
 from callhaskell import *
 
 class Queue(object):
@@ -53,16 +54,8 @@ class GraphState(object):
       return varid
 
   def unify_vars(self, a, b):
-    def replace(v):
-      return a if v == b else v
-
-    def replace_in_factor(fac):
-      (fac_info, args) = fac
-      return (fac_info, map(replace, args))
-
     self.var_replacements[b.id] = a
     del self.vars[b.id]
-    self.factors = map(replace_in_factor, self.factors)
     return a
 
   def unify_values(self, typ, a, b):
@@ -112,10 +105,15 @@ class GraphState(object):
     return rf
 
   def to_JSON(self):
+    def replace_id(varid):
+      if varid.id in self.var_replacements:
+        return self.var_replacements[varid.id].id
+      else:
+        return varid.id
     return {
       'vars': [[varid, expfam] for (varid, expfam) in self.vars.items()],
       'randFuns': [{'argExpFams': aefs, 'resExpFam': ref} for (aefs, ref) in self.rand_funs],
-      'factors': [{'factor': facinfo, 'argVarIds': [a.id for a in args]} for (facinfo, args) in self.factors]
+      'factors': [{'factor': facinfo, 'argVarIds': [replace_id(a) for a in args]} for (facinfo, args) in self.factors]
     }
 
 """
@@ -142,10 +140,10 @@ class DoubleValue(object):
     return Double
 
   def freeze(self, varvals):
-    return varvals[self.varid.id]
+    return varvals[self.varid.id]['value']
 
 gaussian_exp_fam = {'type': 'gaussian'}
-bernoulli = {'type': 'bernoulli'}
+bernoulli_exp_fam = {'type': 'bernoulli'}
 def categorical_exp_fam(n):
   return {'type': 'categorical', 'n': n}
 
@@ -161,7 +159,7 @@ class DoubleClass(object):
     return DoubleValue(vars.dequeue())
 
   def unfreeze(self, state, value):
-    return DoubleValue(state.new_const_var('gaussian', value))
+    return DoubleValue(state.new_const_var(gaussian_exp_fam, value))
 
   def __repr__(self):
     return 'Double'
@@ -176,7 +174,7 @@ class BoolValue(object):
     return Bool
 
   def freeze(self, varvals):
-    return varvals[self.varid.id]
+    return varvals[self.varid.id]['value']
 
 class BoolClass(object):
 
@@ -190,7 +188,7 @@ class BoolClass(object):
[…]

### 2015-02-25 paper/paper.tex
paper and clustering

@@
 
-  Here is an example of a Quipp program to perform clustering:
+  Quipp is implemented as a Python library, so Quipp programs are written as Python programs that have access
+  to special functions.  Here is an example of a Quipp program to perform clustering:
 
@@
     and categorical values.
-    In Quipp, there are more value types than this.  Specifically,
-    in addition to real numbers and categorical types, tuples
-    are also supported.
-    To translate a Quipp program to a factor graph, we
-    map each value to a list of variables, and create factors
-    over all these variables.
+    In Quipp, we have an additonal tuple type, which gets translated
+    to multiple underlying variables in the factor graph.
 
@@
 
+  \subsection{Clustering}
+\begin{verbatim}
+dim = 2
+nclusters = 3
+
+PointType = Vector(dim, Double)
+ClusterType = Categorical(nclusters)
+
+get_cluster = rand_function(ClusterType)
+get_point = rand_function(ClusterType, PointType)
+
+def sample():
+  cluster = get_cluster()
+  return (cluster, get_point(cluster))
+\end{verbatim}
+
+In this example, we cluster 2d points into 3 different clusters.  Given a cluster, the distribution for a point is some independent Gaussian distribution.  This is similar to fuzzy c-means clustering.
+
+  \subsection{Factor Analysis}
+
+\begin{verbatim}
+num_components = 3
+point_dim = 5
+
+ComponentsType = Vector(num_components, Double)
+PointType = Vector(point_dim, Double)
+
+get_point = randFunction(ComponentsType, PointType)
+
+def sample():
+  components = [normal(0, 1) for i in range(num_components)]
+  return (comenents, get_point(components))
+\end{verbatim}
+
+The factor analysis model is very similar to the clustering model.
+The main difference is that we replace the categorical \texttt{ClusterType} type with a vector type.
+This results in the model attempting to find each point as an affine function of a vector of standard normal values.
+
+  \subsection{Hidden Markov model}
+\begin{verbatim}
+num_states = 5
+num_observations = 3
+sequence_length = 10
+
+StateType = Categorical(num_states)
+ObsType = Categorical(num_observations)
+
+trans_fun = rand_function(StateType, StateType)
+obs_fun = rand_function(StateType, ObsType)
+
+def sample():
+  states = [uniform(num_states)]
+  for i in range(sequence_length - 1):
+    states.append(trans_fun(states[-1]))
+  return (states, [obs_fun(s) for s in states])
+\end{verbatim}
+
+In this example, we use the unknown function \texttt{trans\_fun} for state transitions and \texttt{obs\_fun} for observations.
+This means that we will learn both the state transitions and the observation distribution.
+
+\subsection{Latent Dirichlet Allocation}
+
+\begin{verbatim}
+n_classes = 10
+n_words = 10000
+n_words_per_document = 1000
+
+ClassType = Categorical(n_classes)
+WordType = Categorical(n_words)
+
+class_to_word = rand_function(ClassType, WordType)
+
+def sample():
+    which_class = uniform_categorical(n_classes)
+    words = [class_to_word(which_class) for i in range(n_words_per_document)]
+    return (which_class, words)
+\end{verbatim}
+
   \section{Discussion}

### 2015-02-25 src/Main.hs

@@ -58,6 +58,9 @@ decodeTemplate str = makeFactorGraphTemplate
           let facinfo = fac !: "factor" in
           (case facinfo !: "type" of
              "randFun" -> Right (facinfo !: "id")
+             "uniformCategorical" ->
+               let n = facinfo !: "n" in
+               Left $ expFamFactor (categoricalValueExpFam n) [] (replicate (n - 1) 0.0, replicate (n-1) [])
              "constant" -> Left $ constFactor (decodeExpFam (facinfo !: "expFam")) $ case facinfo !: "expFam" !: "type" of
                "gaussian" -> DoubleValue (facinfo !: "value")
                "bernoulli" -> BoolValue (facinfo !: "value")

### 2015-02-25 src/python/examples/hmm.py

@@ -9,7 +9,6 @@ ObsType = Categorical(num_observations)
 trans_fun = rand_function(StateType, StateType)
 obs_fun = rand_function(StateType, ObsType)
 
-
 def sample():
   states = [uniform(num_states)]
   for i in range(9):

### 2015-02-25 src/python/examples/nd_clustering.py

@@ -5,13 +5,10 @@ nclusters = 3
 PointType = Vector(dim, Double)
 ClusterType = Categorical(nclusters)
 
-get_cluster = rand_function(ClusterType)
-
 get_point = rand_function(ClusterType, PointType)
 
 def sample():
-  # cluster = Uniform(nclusters)
-  cluster = get_cluster()
+  cluster = UniformCategorical(nclusters)
   return (cluster, get_point(cluster))
 
 run_example(sample)

### 2015-02-25 src/python/graphbuilder.py

@@ -290,6 +290,11 @@ current_graph_state = GraphState()
 def rand_function(*ts):
   return current_graph_state.rand_function(ts[:-1], ts[-1])
 
+def UniformCategorical(n):
+  v = current_graph_state.new_var(categorical_exp_fam(n))
+  current_graph_state.new_factor({'type': 'uniformCategorical', 'n': n}, [v])
+  return CategoricalValue(v, n)
+
 def conditioned_network(state, typ, sampler, frozen_samps):
   samples = [sampler() for i in range(len(samps))]
   for (latent, s), fs in zip(samples, frozen_samps):

### 2015-02-25 paper/paper.tex
Update intro

@@
 
-  Probabilistic programming languages have made it easier to specify Bayesian
-  models as programs.  Once a model is written as a probabilistic
-  program, it is possible to use generic inference algorithms (such as
-  Metropolis Hastings) to perform inference.
+  By separating model specification and inference, probabilistic programming 
+  has made it easier for non-experts to implement and use probabilistic models.
+  Practitioners frequently have strong intuitions about the {\em structure}
+  of their domain knowledge, such as which latent variables exist and what
+  their causal relations are, and probabilistic programming allows them to encode
+  this knowledge. However, it also requires them to specify the specific parametric
+  shape  and parameterization of any distributions used, and intuitions tend to
+  be much less precise there.
+  We present Quipp, a system that does {\em not} require such specification;
+  instead, random variables and random functions can be left undefined
+  and will automatically be filled in under maximum entropy assumptions
+  based on their types and available datasets.
 
-  In this paper, we discuss a feature of probabilistic programming languages
-  that makes it easier to specify some models.
-  Specifically, we allow the programmer to specify that some random functions
-  are unknown.  The system will fill in these unknown functions with
-  some reasonable default family of distributions based on the types and
-  proceed to infer the parameters to this distribution using the
-  expectation maximization algorithm.
+  Our formalism can concisely express a wide variety of models that machine
+  learning practitioners care about, and we provide an expectation
+  maximization algorithm that can learn the parameters for many of
+  these models with reasonable efficiency. This system makes it easy
+  for non-experts to encode their beliefs about the data and to get
+  predictions based on as few additional assumptions as possible.
 
-  This feature has multiple advantages.  First, it is easier to write
-  models without knowing about the class of models being used.  This should
-  make probabliistic programming more accessible to non-experts.  Secondly,
-  parameter inference is more efficient if specialized algorithms are used
-  rather than the generic algorithms used to infer other random variables
-  (such as Metropolis Hastings).
+  % This feature has multiple advantages.  First, it is easier to write
+  % models without knowing about the class of models being used.  This should
+  % make probabliistic programming more accessible to non-experts.  Secondly,
+  % parameter inference is more efficient if specialized algorithms are used
+  % rather than the generic algorithms used to infer other random variables
+  % (such as Metropolis Hastings).
 
-  We define an example probabilistic programming language with this feature
-  (Quipp) and show how it can be used to write machine learning models
-  concisely.
+  % We define an example probabilistic programming language with this feature
+  % (Quipp) and show how it can be used to write machine learning models
+  % concisely.
 
@@
 
-  Inference in these models is performed using the expectation-maximization
-  algorithm, with alternating steps of inferring latent variables and optimizing
-  parameters.
+  % Inference in these models is performed using the expectation-maximization
+  % algorithm, with alternating steps of inferring latent variables and optimizing
+  % parameters.
 
-  - Motivation for Quipp
-    - Explanation of "unknown functions"
-    - Writing machine learning algorithms as probabilistic programs
-    - Accessibility to non-experts
-    - Comparison to existing probabilistic programming languages
-      - In other languages, use random variables for parameters
-      - Random variables slower because they are updated independently
+  % - Motivation for Quipp
+  %   - Explanation of "unknown functions"
+  %   - Writing machine learning algorithms as probabilistic programs
+  %   - Accessibility to non-experts
+  %   - Comparison to existing probabilistic programming languages
+  %     - In other languages, use random variables for parameters
+  %     - Random variables slower because they are updated independently
+
+  In the following, we first specify the syntax used to write Quipp programs,
+  including the notation for unknown variables and functions.
+  We describe the class of exponential family variables and functions that our system can learn,
+  and present the expectation maximization algorithm used to learn them.
+  We then demonstrate the expressiveness of our language, and the broad
+  applicability of our algorithm, by writing some of the most common machine learning models
+  in Quipp: clustering, factor analysis, a Hidden Markov model, Latent Dirichlet Allocation, and
+  a neural net.
 
@@
 
-    Let $\psi_x$ be a matrix such that $\psi \phi(x)$ is a vector of the
+    Let $\psi_x$ be a matrix such that $\psi \phi(x)$ is a vector x of the
     features of $x$.  Then

### 2015-02-25 paper/paper.tex
more paper

--- 2015-02-25 paper/paper.tex
@@
   \begin{verbatim}
-  PointType = Vector(2, Double)
-  ClusterType = Categorical(3)
+  dim = 2
+  nclusters = 3
+  PointType = Vector(dim, Double)
+  ClusterType = Categorical(nclusters)
 
-  get_cluster = rand_function(ClusterType)
   get_point = rand_function(ClusterType, PointType)
@@
   def sample():
-    cluster = get_cluster()
+    cluster = uniform_categorical(nclusters)
     return (cluster, get_point(cluster))
@@
   Notice that we declared two types (\texttt{PointType} and \texttt{ClusterType})
-  and two random functions (\texttt{get\_cluster} and \texttt{get\_point}).
+  and one random function \texttt{get\_point}).
   Type annotations are necessary for random functions.  The type
@@
   and the type \texttt{Categorical(3)} expresses the fact that there are 3
-  possible clusters (so a cluster may be either 0, 1, or 2).  We do not know
-  the distribution over clusters or over points given a cluster.
-  We will fill in the \texttt{get\_cluster} function with
-  a learned categorical distribution, and we will fill in the
+  possible clusters (so a cluster may be either 0, 1, or 2).  We assume that
+  the disttribution over clusters is uniform, but we do not know the distribution
+  of points in each cluster.
+  We will fill in the
   \texttt{get\_point} function with a learned unknown linear function with noise.
@@
 
-    For unknown functions, the class of functions used is a kind
+    For unknown functions, the class of random functions used is a kind
     of generalized linear model.  We assume that the result
-    is distributed from some exponential family.  We label
+    is distributed from some exponential family whose natural
+    parameters are determined from the arguments.  We label
     some subset of the sufficient statistics as \emph{features}.  The natural
@@
 
+We randomly generated parameters for this example 100 times, and each time took 100 samples and then ran 10 EM iterations.
+The accuracy is defined as the maximum percentage of points assigned to the correct cluster, for any permutation of clusters.
+On average, accuracy increased in each EM iteration, as shown it this graph:
+
+\begin{center}
+\includegraphics[scale=0.5]{cluster_accuracy.png}
+\end{center}
+
+\subsection{Naive Bayes}
+
+\begin{verbatim}
+nfeatures = 10
+FeaturesType = Vector(nfeatures, Bool)
+
+get_class = rand_function(Unit, Bool)
+
+class_1_features = rand_function(Unit, FeaturesType)
+class_2_features = rand_function(Unit, FeaturesType)
+
+def sample():
+  which_class = get_class()
+  if which_class:
+    features = class_1_features()
+  else:
+    features = class_2_features()
+  return (which_class, features)
+\end{verbatim}
+
+The naive Bayes model is similar to the clustering model.  We have two classes and a feature
+distribution for each.  Since each feature is boolean, we will learn
+a different categorical distribution for each class.
+
+(figure should show average classification accuracy)
+
   \subsection{Factor Analysis}
@@
 
-get_point = randFunction(ComponentsType, PointType)
+get_point = rand_function(ComponentsType, PointType)
 
@@
 
+(figure should show accuracy by EM iteration.  Accuracy can be measured by distances between different factor coefficients, when ordered according to some permutation)
   \subsection{Hidden Markov model}
@@
 n_classes = 10
-n_words = 10000
+n_words = 100
 n_words_per_document = 1000
@@
 
+In this example, we use the unknown function \texttt{class\_to\_word} to map classes to word distributions.
+Note that each column of the matrix of parameters for \texttt{class\_to\_word} will represent a categorical distribution over words, and there will be one column for each class.
+
+(figure should show accuracy over time.  Accuracy can be measured as distance between the learned categorical distributions, for some permutation of classes)
+
+\subsection{Neural Network}
+
+\begin{verbatim}
+input_dim = 100
+hidden_dim = 20
+output_dim = 2
+
+InputType = Categorical(input_dim)
+HiddenType = Categorical(hidden_dim)
+OutputType = Categorical(output_dim)
+
+input_to_hidden = rand_function(InputType, HiddenType)
+hidden_to_output = rand_function(HiddenType, OutputType)
+
+def sample(input_layer):
+  hidden_layer = input_to_hidden(input_layer)
+  output_layer = hidden_to_output(hidden_layer)
+  return ((), output_layer)
+\end{verbatim}
+
+\subsection{A more complex model}
+
+TODO: come up with a more complex model.  Show how this model can be iteratively improved by adding information.  It would make sense for this to be some sort of data science type
+
   \section{Discussion}

### 2015-02-25 src/python/examples/nd_clustering.py

@@ -1,14 +1,16 @@
 from quipp import *
 
-dim = 2
-nclusters = 3
-PointType = Vector(dim, Double)
-ClusterType = Categorical(nclusters)
+def run():
+  dim = 2
+  nclusters = 3
+  PointType = Vector(dim, Double)
+  ClusterType = Categorical(nclusters)
 
-get_point = rand_function(ClusterType, PointType)
+  get_point = rand_function(ClusterType, PointType)
 
-def sample():
-  cluster = UniformCategorical(nclusters)
-  return (cluster, get_point(cluster))
+  def sample():
+    cluster = UniformCategorical(nclusters)
+    return (cluster, get_point(cluster))
+  return sample
 
-run_example(sample)
+run_example(run)

### 2015-02-25 src/python/graphbuilder.py

@@ -1,4 +1,5 @@
 import math
+import itertools
 from callhaskell import *
 
 class Queue(object):
@@ -325,18 +326,66 @@ def translate_params_for_fn(params):
     factors = [params[0][0]] + params[1][0]
     return (variance, [f*variance for f in factors])
 
+def params_to_cluster_centers(params):
+  d = dict(params)
+  cluster_centers = []
+  for i in d:
+    ps = d[i]
+    variance = -1.0 / (2 * ps[0][1])
+    factors = [ps[0][0]] + ps[1][0]
+    scaled_factors = [f*variance for f in factors]
+    centers = [scaled_factors[0]] + [x + scaled_factors[0] for x in scaled_factors[1:]]
+    cluster_centers.append(centers)
+  return zip(*cluster_centers)
+
+def cluster_centers_error(cs1, cs2):
+  errs = []
+  def tup_dist(t1, t2):
+    return sum((a-b)**2 for (a, b) in zip(t1, t2))
+  for cs in itertools.permutations(cs1):
+    errs.append(sum(map(tup_dist, cs, cs2)))
+  return min(errs)
+
+def cluster_assignment_accuracy(cs1, cs2):
+  accuracies = []
+  for perm in itertools.permutations(range(3)):
+    accuracies.append(float(len([() for (a, b) in zip(cs1, cs2) if a == perm[b]])) / len(cs1))
+  return max(accuracies)
+
+
+
 def translate_params(params):
   return [(x, translate_params_for_fn(y)) for (x, y) in params]
 
-def run_example(sampler):
-  n = 1000
-  samples = [sampler() for i in range(n)]
-  templ = current_graph_state.to_JSON()
-  rand_params = hs_rand_template_params(templ)
-  varvals = state_to_varvals(hs_sample_bayes_net(templ, rand_params))
-  # TODO freeze only second value
-  frozen_samples = [freeze_value(samp, varvals) for samp in samples]
-  state_params_list = infer_parameters_from_samples(current_graph_state, samples, [x[1] for x in frozen_samples])
-  print translate_params(rand_params)
-  print translate_params(state_params_list[-1][1])
-  print str(state_params_list[-2][0]) == str(state_params_list[-1][0])
+
+def mean(xs):
+  return sum(xs) / len(xs)
+
+def run_example(run):
+  global current_graph_state
+  n = 100
+  accs = []
+  for i in range(100):
+    current_graph_state = GraphState()
+    sampler = run()
+    samples = [sampler() for i in range(n)]
[…]

### 2015-02-25 paper/paper.tex
paper

@@
 \setlength{\emergencystretch}{10em}
+\usepackage{proceed2}
 
@@
 
-  \section{Introduction}
+  \section{INTRODUCTION}
 
@@
   applicability of our algorithm, by writing some of the most common machine learning models
-  in Quipp: clustering, factor analysis, a Hidden Markov model, Latent Dirichlet Allocation, and
+  in Quipp: clustering, naive Bayes, factor analysis, a Hidden Markov model, Latent Dirichlet Allocation, and
   a neural net.
 
-  \section{Syntax}
+  \section{SYNTAX}
 
   Quipp is implemented as a Python library, so Quipp programs are written as Python programs that have access
-  to special functions.  Here is an example of a Quipp program to perform clustering:
+  to special functions.  Here is an example of a Quipp program to cluster 2d points into 3 clusters:
 
@@
   possible clusters (so a cluster may be either 0, 1, or 2).  We assume that
-  the disttribution over clusters is uniform, but we do not know the distribution
+  the distribution over clusters is uniform, but we do not know the distribution
   of points in each cluster.
   We will fill in the
-  \texttt{get\_point} function with a learned unknown linear function with noise.
+  \texttt{get\_point} function with a learned function that will take a random
+  sample from the given cluster.
 
-  \section{Class of Distributions}
+  \section{FAMILY OF DISTRIBUTIONS}
 
-    For unknown functions, the class of random functions used is a kind
+  For unknown functions, the family of random functions used is a kind
     of generalized linear model.  We assume that the result
     is distributed from some exponential family whose natural
-    parameters are determined from the arguments.  We label
-    some subset of the sufficient statistics as \emph{features}.  The natural
+    parameters are determined from the arguments:
+
+    $$p_{\eta}(y | x) = \exp\left(\eta(x)^T \phi(y) - g(\eta(x))\right)$$
+
+    Here, $\eta$ is the natural parameter, $\phi(y)$ is a vector of $y$'s sufficient statistics,
+    and $g$ is the log partition function.
+
+    To determine $\eta$ as a function of $x$, we label
+    some subset of the sufficient statistics of both $x$ and $y$ as \emph{features}.  The natural
     parameters corresponding to non-features are constant, while natural
     parameters corresponding to features are determined as an affine
-    function of the features of the arguments.
+    function of the features of the arguments.  Features are the same as sufficient
+    statistics for the categorical distribution, but while both $X$ and $X^2$ are
+    sufficient statistics for the Gaussian distribution, only $X$ is a feature.
 
-    Let $\psi_x$ be a matrix such that $\psi \phi(x)$ is a vector x of the
-    features of $x$.  Then
+    Let $\psi(x)$ be the features of $x$.  Then
+    $$p_{\mathbf{N}}(y | x) = \exp\left(\begin{bmatrix} 1 \\ \psi(x) \end{bmatrix} ^T \mathbf{N} \phi(y) - g\left(\mathbf{N}^T \begin{bmatrix} 1 \\ \psi(x) \end{bmatrix}\right)\right)$$
 
-    $$p_{\mathbf{N}}(y | x) = \exp(\psi(x)^T N \phi(y) - g(N^T \psi(X)))$$
+    where $\mathbf{N}$ is a matrix containing our parameters.  It must have 0 for entries not in the first
+    row corresponding to sufficient statistics of $y$ that are not features.  This ensures that only the
+    natural parameters for features are affected by $x$.
 
     Given $(x, y)$ samples, parameter estimation to maximize log probability is a convex
-    problem:
+    problem because the log probability function is concave:
+    $$\log p_{\mathbf{N}}(y | x) = \begin{bmatrix} 1 \\ \psi(x) \end{bmatrix} ^T \mathbf{N} \phi(y) - g\left(\mathbf{N}^T \begin{bmatrix} 1 \\ \psi(x) \end{bmatrix}\right)$$
 
-    \begin{align*}
-      \log p_{\mathbf{N}}(y | x) = \psi(x)^T N \phi(y) - g(N^T \psi(X))
-    \end{align*}
-    since the log probability function is concave when $g$ is convex (as it is
-    for any exponential family distribution).
+    This relies on the fact that $g$ is convex, but this is true in general for any exponential family distribution.
+    Since the problem is convex, it is possible to use Newton's method to optimize the parameters.  Although
+    the only exponential family distributions we use in this paper are the categorical and Gaussian distributions,
+    we can use the same algorithms for other exponential families, such as the Poisson and gamma distributions.
 
-    Therefore, it is possible to use Newton's method to optimize the parameters.
+  \section{INFERENCE}
 
-  \section{Inference}
-
-    To perform inference of both latent variables and parameters, we first
+    To infer both latent variables and parameters, we first
     translate the program to a factor graph.  Next, we run the
     expectation maximization algorithm on this graph, iterating stages of
-    estimating latent variables using Metropolis Hastings and parameter
-    inference using Newton's method.
+    estimating latent variables using Metropolis Hastings and inferring
+    parameters using Newton's method.
 
@@
     In Quipp, we have an additonal tuple type, which gets translated
-    to multiple underlying variables in the factor graph.
+    to multiple underlying variables in the factor graph.  In the
+    clustering example, we would represent each point by 2 real number variables,
+    and each cluster assignment as a single categorical variable (with 3 alternatives).
 
-    There are a few different types of factors:
-    \begin{enumerate}
-      \item Factors representing known distributions
-      \item Factors representing unknown distribution
-      \item Factors asserting that a value is equal to some constant
-    \end{enumerate}
+    Factors can represent known distributions,
+    unknown distributions, or observations.
+    In the clustering example, we will have, for each cluster assignment variable,
+    a factor representing the fact that it comes from a (known) uniform distribution.
+    For each real number in a data point, we will have a factor connecting it to
+    the cluster assignment for the point, which will represent the unknown distribution.
+    This factor depends on the specific values of the parameters for the $\texttt{get\_point}$ function.
+    Also, since data points are observed, they will have constant factors attached to them
+    representing these observations.
+
+    \begin{figure}[h]
+    \begin{center}
+    \includegraphics[scale=0.2]{cluster_diagram.png}
+    \end{center}
+    \caption{Translating clustering model to factor graph.  TODO fix variable names} 
+    \end{figure}
 
@@
     $f(x_1, x_2, ..., x_n)$ and each $1 \leq i \leq n$, we send a message
-    to variable $x_i$ that approximates $x_i \mapsto f(x_1, x_2, ..., x_n)$.
+    to variable $x_i$ that approximates $x_i \mapsto \log f(x_1, x_2, ..., x_n)$ as
+    an affine function as $\phi(x_i)$.  The proposal distribution is then the
+    product of these messages, which will approximate the local distribution of
+    the variable.
 
-  \section{Examples}
+  \section{EXAMPLES}
 
-  \subsection{Clustering}
+  \subsection{CLUSTERING}
+{\small
 \begin{verbatim}
@@
 \end{verbatim}
+}
 
@@
 \begin{center}
-\includegraphics[scale=0.5]{cluster_accuracy.png}
+\includegraphics[scale=0.4]{cluster_accuracy.png}
 \end{center}
 
-\subsection{Naive Bayes}
+\subsection{NAIVE BAYES}
 
+{\small
 \begin{verbatim}
@@
 
-  \subsection{Factor Analysis}
-
+  \subsection{FACTOR ANALYSIS}
+{\small
 \begin{verbatim}
@@
 def sample():
-  components = [normal(0, 1) for i in range(num_components)]
+  components = [normal(0, 1)
+                for i in range(num_components)]
   return (comenents, get_point(components))
 \end{verbatim}
+}
 
@@
 (figure should show accuracy by EM iteration.  Accuracy can be measured by distances between different factor coefficients, when ordered according to some permutation)
-  \subsection{Hidden Markov model}
+  \subsection{HIDDEN MARKOV MODEL}
+{\small
 \begin{verbatim}
@@
 
-\subsection{Latent Dirichlet Allocation}
-
+\subsection{LATENT DIRICHLET ALLOCATION}
+{\small
 \begin{verbatim}
@@
     which_class = uniform_categorical(n_classes)
-    words = [class_to_word(which_class) for i in range(n_words_per_document)]
+    words = [class_to_word(which_class)
+             for i in range(n_words_per_document)]
     return (which_class, words)
 \end{verbatim}
+}
 
@@
 
-\subsection{Neural Network}
+\subsection{NEURAL NETWORK}
 
+{\small
 \begin{verbatim}
@@
 \end{verbatim}
+}
 
-\subsection{A more complex model}
+\subsection{A MORE COMPLEX MODEL}
 
-TODO: come up with a more complex model.  Show how this model can be iteratively improved by adding information.  It would make sense for this to be some sort of data science type
+TODO: if there is time, we should put a more complex data science type example, where we add dependencies and show change in accuracy as we add/remove assumptions.
 
-  \section{Discussion}
+  \section{DISCUSSION}
 
-  - Comments about results
-  - Future directions
-    - Unbounded models
-    - Recursive algebraic data types
+  We have found that it is possible to write many useful machine learning models as Quipp programs and then use generic algorithms for inference.
+Furthermore, performance is <???>.
+This should make it much easier for non-experts to write useful machine learning models.
+
+  In the future, it will be useful to support unbounded recursion in models (as in Church).
+Furthermore, it will also be useful to expand the set of types supported.
+It is possible to define reasonable default distributions for non-recursive algebraic data types, and it may also be possible to define them for recursive algebraic data types using catamorphisms.
 

### 2015-02-28 src/Main.hs
updates

@@ -61,10 +61,15 @@ decodeTemplate str = makeFactorGraphTemplate
              "uniformCategorical" ->
                let n = facinfo !: "n" in
                Left $ expFamFactor (categoricalValueExpFam n) [] (replicate (n - 1) 0.0, replicate (n-1) [])
+             "normal" ->
+               let mean = facinfo !: "mean"
+                   stdev = facinfo !: "stdev"
+              in Left $ expFamFactor gaussianValueExpFam [] ([mean / stdev^2, -1 / (2 * stdev^2)], [[]])
              "constant" -> Left $ constFactor (decodeExpFam (facinfo !: "expFam")) $ case facinfo !: "expFam" !: "type" of
                "gaussian" -> DoubleValue (facinfo !: "value")
                "bernoulli" -> BoolValue (facinfo !: "value")
                "categorical" -> CategoricalValue (facinfo !: "value")
+             other -> error $ "Unknown factor type: " ++ other
 
            , fac !: "argVarIds")
 

### 2015-02-28 src/python/examples/factor_analysis.py

@@ -1,15 +1,18 @@
 from quipp import *
 
-num_components = 3
-point_dim = 5
+def run():
+  num_components = 2
+  point_dim = 5
 
-ComponentsType = Vector(num_components, Double)
-PointType = Vector(point_dim, Double)
+  ComponentsType = Vector(num_components, Double)
+  PointType = Vector(point_dim, Double)
 
-get_point = randFunction(ComponentsType, PointType)
+  get_point = rand_function(ComponentsType, PointType)
 
-def sample():
-  components = [normal(0, 1) for i in range(num_components)]
-  return (comenents, get_point(components))
+  def sample():
+    components = [normal(0, 1) for i in range(num_components)]
+    return (components, get_point(components))
 
-run_example(sample)
+  return sample
+
+run_example(run)

### 2015-02-28 src/python/examples/nd_clustering.py

@@ -9,8 +9,8 @@ def run():
   get_point = rand_function(ClusterType, PointType)
 
   def sample():
-    cluster = UniformCategorical(nclusters)
+    cluster = uniform_categorical(nclusters)
     return (cluster, get_point(cluster))
   return sample
 
-run_example(run)
+run_clustering_example(run)

### 2015-02-28 src/python/graphbuilder.py

@@ -1,4 +1,6 @@
 import math
+import numpy as np
+from numpy import linalg
 import itertools
 from callhaskell import *
 
@@ -268,7 +270,7 @@ class Categorical(object):
 def get_type(value):
   if hasattr(value, 'get_type'):
     return value.get_type()
-  elif isinstance(value, tuple):
+  elif isinstance(value, (tuple, list)):
     return Tuple(*map(get_type, value))
   else:
     raise Exception('Unknown value type ' + str(type(value)) + ', value ' + str(value))
@@ -276,12 +278,13 @@ def get_type(value):
 def freeze_value(value, varvals):
   if hasattr(value, 'freeze'):
     return value.freeze(varvals)
-  elif isinstance(value, tuple):
+  elif isinstance(value, (tuple, list)):
     return tuple([freeze_value(v, varvals) for v in value])
   else:
     raise Exception('Unknown value type ' + str(type(value)) + ', value ' + str(value))
 
 def Vector(n, typ):
+  print 'vector', n, typ, Tuple(*([typ]*n))
   return Tuple(*([typ]*n))
 
 Unit = Tuple()
@@ -291,10 +294,16 @@ current_graph_state = GraphState()
 def rand_function(*ts):
   return current_graph_state.rand_function(ts[:-1], ts[-1])
 
-def UniformCategorical(n):
+def uniform_categorical(n):
   v = current_graph_state.new_var(categorical_exp_fam(n))
   current_graph_state.new_factor({'type': 'uniformCategorical', 'n': n}, [v])
   return CategoricalValue(v, n)
+
+def normal(mean, stdev):
+  v = current_graph_state.new_var(gaussian_exp_fam)
+  current_graph_state.new_factor({'type': 'normal', 'mean': mean, 'stdev': stdev}, [v])
+  return DoubleValue(v)
+
 
 def conditioned_network(state, typ, sampler, frozen_samps):
   samples = [sampler() for i in range(len(samps))]
@@ -309,12 +318,11 @@ def infer_parameters_from_samples(graph_state, samples, frozen_samples):
     graph_state.unify_values(typ, s[1], typ.unfreeze(graph_state, f))
   templ = graph_state.to_JSON()
   (state, params) = hs_init_em(templ)
-  state_params_list = [(state, params)]
+  yield (state, params)
[…]

### 2015-03-01 src/Main.hs
factor analysis kinda works!

@@ -124,7 +124,7 @@ pyInitEM templ = do
 pyStepEM :: FactorGraphTemplate Value -> FST -> IO FST
 pyStepEM templ (state, params) = sampleRVar $ do
   let factorGraph = instantiateTemplate templ params
-  newStates <- sampleRVarTWith (\(Just x) -> return x) $ iterateM 10 (stepMH factorGraph) state
+  newStates <- sampleRVarTWith (\(Just x) -> return x) $ iterateM 20 (stepMH factorGraph) state
   let params' = updateTemplateParams templ params [(1.0, s) | s <- takeEvery 1 (tail newStates)]
   return (last newStates, params')
 

### 2015-03-01 src/Quipp/ExpFam.hs

@@ -4,7 +4,7 @@ module Quipp.ExpFam (ExpFam(ExpFam, expFamName, expFamD, expFamSufStat, expFamG,
                      expFamFeaturesD, expFamSufStatToFeatures, expFamFeaturesToSufStat,
                      expFamSufStatToLikelihood, expFamCrossEntropy,
                      getNatParam,
-                     Params,
+                     Params, mapParams,
                      promoteExpFam, expFamLogProbability, expFamMLE, expFamMH,
                      Likelihood(KnownValue, NatParam), promoteLikelihood, negInfinity,
                      likelihoodLogProbability,
@@ -115,6 +115,8 @@ mhNewtonMethod f x = iterateRVar (mhNewtonMethodStep f) x
 ratNum = (fromRational :: Rational -> Double) . toRational
 
 type Params m = ([m], Matrix m)
+
+mapParams f (xs, ys) = (map f xs, map (map f) ys)
 
 paramsToVector :: Params m -> [m]
 paramsToVector (base, weights) = base ++ concat weights

### 2015-03-01 src/Quipp/Factor.hs

@@ -1,4 +1,4 @@
-{-# LANGUAGE TupleSections #-}
+{-# LANGUAGE TupleSections, RankNTypes, NoMonomorphismRestriction, ViewPatterns #-}
 
 module Quipp.Factor (
   Factor(Factor, factorExpFams, factorLogValue, factorNatParam, factorBayesNetOutput),
@@ -22,7 +22,7 @@ import Data.Map (Map, (!))
 import qualified Data.Map as Map
 import Data.Maybe (fromJust)
 import Data.Random (RVar, normal)
-import Numeric.AD (grad)
+import Numeric.AD (grad, diff)
 
 import Quipp.Value
 import Quipp.ExpFam
@@ -71,6 +71,13 @@ constFactor ss x = Factor {
 
 
 
+getFeatureMessage :: ExpFam a -> (forall b. RealFloat b => [b] -> b) -> [Double] -> [Double]
+getFeatureMessage (expFamName -> "gaussian") f [x] = [d - d2 * x, d2 / 2]
+  where f' = f . (:[])
+        d = diff f' x
+        d2 = diff (diff f') x
+getFeatureMessage (expFamName -> "categorical") f x = grad f x
+
 -- TODO: something is fishy here.  Things get flipped when they shouldn't.
 expFamFactor :: ExpFam v -> [ExpFam v] -> Params Double -> Factor v
 expFamFactor ef argExpFams eta@(etaBase, etaWeights) =
@@ -87,8 +94,15 @@ expFamFactor ef argExpFams eta@(etaBase, etaWeights) =
           let gradProbNp = grad (\np -> dotProduct np (map fromDouble ss) - expFamG ef np) $ getNatParam ef eta feats
               minFeatureIndex = sum $ map expFamFeaturesD $ take (n-1) argExpFams
               thisArgDim = expFamFeaturesD (argExpFams !! (n-1))
-              relevantWeights = map (take thisArgDim . drop minFeatureIndex) etaWeights
-          in expFamFeaturesToSufStat (argExpFams !! (n-1)) $ matMulByVector (transpose relevantWeights) gradProbNp
+              probFromFeats :: RealFloat a => [a] -> a
+              probFromFeats feats' =
+                expFamLogProbability ef (mapParams fromDouble eta)
+                  (map fromDouble (take minFeatureIndex feats) ++ feats' ++ map fromDouble (drop (minFeatureIndex + thisArgDim) feats)) (map fromDouble ss)
+
[…]

### 2015-03-01 src/python/graphbuilder.py

@@ -319,10 +319,12 @@ def infer_parameters_from_samples(graph_state, samples, frozen_samples):
   templ = graph_state.to_JSON()
   (state, params) = hs_init_em(templ)
   yield (state, params)
-  for i in range(10):
+  i = 0
+  while True:
     print 'iter', i
     (state, params) = hs_step_em(templ, state, params)
     yield (state, params)
+    i += 1
 
 def translate_params_for_fn(params):
   if len(params[1][0]) == 0:
@@ -401,21 +403,24 @@ def run_clustering_example(run):
 
 def params_to_matrix(params):
   coords = []
+  component_variances = []
   for (i, ((base_n1, n2), (lin,))) in params:
+    component_variances.append(-1.0 / (2 * n2))
     coords.append([-l/(2 * n2) for l in [base_n1] + lin])
-  return np.matrix(coords)
+  return component_variances, np.matrix(coords)
 
-def matrix_to_gaussian(mat):
+def matrix_to_gaussian(variances_mat):
+  variances, mat = variances_mat
   mean = mat[:,0]
   a = mat[:, 1:]
-  return (mean, a * a.T)
+  return (mean, a * a.T + np.diag(variances))
 
 def gaussian_kl(p, q):
   (pm, pv) = p
   (qm, qv) = q
   n = pm.shape[0]
   assert pv.shape[0] == n == qv.shape[0]
-  return 0.5 * (np.trace(linalg.inv(qv) * pv) + (qm - pm).T * linalg.inv(pv) * (qm - pm) - n + np.slogdet(qv)[1] - np.slogdet(pv)[1])
+  return 0.5 * (np.trace(linalg.inv(qv) * pv) + ((qm - pm).T * linalg.inv(qv) * (qm - pm)).item((0,0)) - n + linalg.slogdet(qv)[1] - linalg.slogdet(pv)[1])
 
 def rotation_invariant_dist(A, B):
   # min_R ||AR - B||^2
@@ -440,7 +445,7 @@ def rotation_invariant_dist(A, B):
 
 def run_example(run):
   global current_graph_state
-  n = 100
+  n = 200
   accs = []
   for i in range(1):
     current_graph_state = GraphState()
@@ -450,7 +455,6 @@ def run_example(run):
     rand_params = hs_rand_template_params(templ)
     rand_mat = params_to_matrix(rand_params)
     print rand_mat
-    print rotation_invariant_dist(rand_mat, rand_mat)
     varvals = state_to_varvals(hs_sample_bayes_net(templ, rand_params))
[…]

### 2015-03-01 src/python/examples/factor_analysis.py
rename

@@ -15,4 +15,4 @@ def run():
 
   return sample
 
-run_example(run)
+run_factor_analysis_example(run)

### 2015-03-01 src/python/graphbuilder.py

@@ -443,7 +443,7 @@ def rotation_invariant_dist(A, B):
 
 # IDEA: compute Gaussian from factors, KL divergence!
 
-def run_example(run):
+def run_factor_analysis_example(run):
   global current_graph_state
   n = 200
   accs = []

### 2015-03-01 src/Quipp/ExpFam.hs
regularization

@@ -141,6 +141,20 @@ groupSamplesByFeatures samps =
   in [(totWeight, snd3 (head g), map sum $ transpose [scaleVec (w / totWeight) s | (w, _, s) <- g])
       | g <- grouped, let totWeight = sum $ map fst3 g]
 
+sampleVariances :: [(Double, [Double], [Double])] -> ([Double], [Double])
+sampleVariances xs = (var snd3, var thd3)
+  where totWeight = sum (map fst3 xs)
+        minVar = 0.00001
+        -- normalize handles NaN too
+        normalize x = if x >= minVar then x else minVar
+        var f = map normalize $ zipWith (-) (map (^2) (ev f id)) (ev f (^2))
+        ev f g = foldl1 (zipWith (+)) [scaleVec (fst3 x / totWeight) (map g $ f x) | x <- xs]
+
+-- normalize entry for feature a, suf stat b, by stdev(b) / stdev(a)
+-- that is, regularization factor is proportional to stdev(a) / stdev(b)
+  
+  
+
 -- Special case for Gaussian (linear regression).  I have confirmed that
 -- this is equivalent.  Note that this does not work for VMP or for
 -- samples with weight not equal to 1.
@@ -157,9 +171,12 @@ expFamMLE fam samples etaStart | expFamName fam == "gaussian" =
 
 expFamMLE fam samples etaStart = -- trace ("\nexpFamMLE " ++ show samples) $
   let samples' = groupSamplesByFeatures samples
+      (xvars, yvars) = sampleVariances samples
       f :: RealFloat m => [m] -> m
-      f eta = sum [fromDouble weight * expFamLogProbability fam params (map fromDouble exs) (map fromDouble ys) | (weight, exs, {- varxs, -} ys) <- samples']
+      f eta = -baseReg (fst params) - featReg (snd params) + sum [fromDouble weight * expFamLogProbability fam params (map fromDouble exs) (map fromDouble ys) | (weight, exs, {- varxs, -} ys) <- samples']
         where params = vectorToParams fam eta
+              baseReg ns = squareMagnitude $ zipWith (/) ns (map (fromDouble . sqrt) yvars)
+              featReg ns = squareMagnitude $ zipWith (*) (concat ns) [sqrt (fromDouble xv / fromDouble yv) | yv <- expFamSufStatToFeatures fam yvars, xv <- xvars]
[…]

### 2015-03-01 src/Quipp/Factor.hs

@@ -17,7 +17,7 @@ import Control.Monad (liftM, replicateM)
 import Control.Monad.State.Class (get, put)
 import Control.Monad.State.Lazy (evalState)
 import Debug.Trace
-import Data.List (elemIndex)
+import Data.List (elemIndex, isPrefixOf)
 import Data.Map (Map, (!))
 import qualified Data.Map as Map
 import Data.Maybe (fromJust)
@@ -76,7 +76,8 @@ getFeatureMessage (expFamName -> "gaussian") f [x] = [d - d2 * x, d2 / 2]
   where f' = f . (:[])
         d = diff f' x
         d2 = diff (diff f') x
-getFeatureMessage (expFamName -> "categorical") f x = grad f x
+getFeatureMessage ef f x | "categorical" `isPrefixOf` expFamName ef = grad f x
+getFeatureMessage ef f x = error $ "Bad expfam and features: " ++ expFamName ef ++ ", " ++ show x
 
 -- TODO: something is fishy here.  Things get flipped when they shouldn't.
 expFamFactor :: ExpFam v -> [ExpFam v] -> Params Double -> Factor v

### 2015-03-01 src/Quipp/Util.hs

@@ -109,6 +109,9 @@ transpose xs = if maximum (map length xs) == 0 then [] else map head xs : transp
 dotProduct :: Num a => [a] -> [a] -> a
 dotProduct x y = sum (zipWith (*) x y)
 
+squareMagnitude :: Num a => [a] -> a
+squareMagnitude x = dotProduct x x
+
 matMulByVector :: Num a => Matrix a -> [a] -> [a]
 matMulByVector m v = map (dotProduct v) m
 

### 2015-03-01 src/python/examples/hmm.py

@@ -1,18 +1,21 @@
 from quipp import *
 
-num_states = 5
-num_observations = 3
 
-StateType = Categorical(num_states)
-ObsType = Categorical(num_observations)
+def run():
+  num_states = 3
+  num_observations = 3
 
-trans_fun = rand_function(StateType, StateType)
-obs_fun = rand_function(StateType, ObsType)
+  StateType = Categorical(num_states)
+  ObsType = Categorical(num_observations)
 
-def sample():
-  states = [uniform(num_states)]
-  for i in range(9):
-    states.append(trans_fun(states[-1]))
-  return (states, [obs_fun(s) for s in states])
+  trans_fun = rand_function(StateType, StateType)
+  obs_fun = rand_function(StateType, ObsType)
 
-run_example(sample)
+  def sample():
+    states = [uniform_categorical(num_states)]
+    for i in range(9):
+      states.append(trans_fun(states[-1]))
+    return (states, [obs_fun(s) for s in states])
+  return sample
+
+run_hmm_example(run)

### 2015-03-01 src/python/graphbuilder.py

@@ -381,6 +381,7 @@ def run_clustering_example(run):
     samples = [sampler() for i in range(n)]
     templ = current_graph_state.to_JSON()
     rand_params = hs_rand_template_params(templ)
+    print hs_sample_bayes_net(templ, rand_params)
     varvals = state_to_varvals(hs_sample_bayes_net(templ, rand_params))
     frozen_samples = [freeze_value(samp, varvals) for samp in samples]
     true_latents = [x[0] for x in frozen_samples]
@@ -479,6 +480,40 @@ def run_factor_analysis_example(run):
         prev_state_latents = state_latents
         # acc = cluster_assignment_accuracy(true_latents, state_latents)
         # iter_accs.append(acc)
+      j += 1
+
+    # accs.append(iter_accs)
+  # print map(mean, zip(*accs))
+
+def run_hmm_example(run):
+  global current_graph_state
+  n = 100
+  accs = []
+  for i in range(1):
+    current_graph_state = GraphState()
+    sampler = run()
+    samples = [sampler() for i in range(n)]
+    templ = current_graph_state.to_JSON()
+    rand_params = hs_rand_template_params(templ)
+    print rand_params
+    # rand_mat = params_to_matrix(rand_params)
+    varvals = state_to_varvals(hs_sample_bayes_net(templ, rand_params))
+    frozen_samples = [freeze_value(samp, varvals) for samp in samples]
+    true_latents = [x[0] for x in frozen_samples]
+    # print true_latents
+    state_params_list = infer_parameters_from_samples(current_graph_state, samples, [x[1] for x in frozen_samples])
+    # rand_cs = params_to_cluster_centers(rand_params)
+    iter_accs = []
+    j = 0
+    prev_state_latents = None
+    for (state, params) in state_params_list:
+      # guess_mat = params_to_matrix(params)
+      # cs = params_to_cluster_centers(params)
+      if j > 1:
+        print params
+        varvals = state_to_varvals(state)
+        state_latents = [freeze_value(samp[0], varvals) for samp in samples]
+        prev_state_latents = state_latents
       j += 1
 
     # accs.append(iter_accs)

### 2015-03-02 src/Main.hs
variational message passing and scoring works!

@@ -1,13 +1,15 @@
-{-# LANGUAGE FlexibleInstances #-}
+{-# LANGUAGE FlexibleInstances, NoMonomorphismRestriction, ScopedTypeVariables #-}
 module Main where
 
 import Debug.Trace (trace, traceShow)
-import Control.Monad (liftM)
+import Control.Monad (liftM, replicateM)
 import Control.Applicative ((<$>), (<*>))
 import Data.Map (Map, (!))
 import qualified Data.Map as Map
 import System.Environment (getArgs)
 import qualified Text.JSON as JSON
+import Data.Maybe (fromJust)
+import Data.Random (RVarT, RVar)
 
 import Quipp.BayesNet
 import Quipp.ExpFam
@@ -44,40 +46,34 @@ infixl 9 !:
     JSON.Error str -> error $ "Error looking up " ++ key ++ " in " ++ show obj ++ ": " ++ str
 other !: key = error ("Not object: " ++ show other)
 
-decodeTemplate :: String -> FactorGraphTemplate Value
-decodeTemplate str = makeFactorGraphTemplate
-    (map convVar (m !: "vars"))
-    (zip1_2 [0..] $ map convRandFun (m !: "randFuns"))
-    (zip1_2 [0..] $ map convFactor (m !: "factors"))
-  where m = decodeJSON str
-        convVar (i, ef) = (i, decodeExpFam ef)
-        convRandFun rf =
-          (decodeExpFam (rf !: "resExpFam"),
-           map decodeExpFam (rf !: "argExpFams"))
-        convFactor fac =
-          let facinfo = fac !: "factor" in
-          (case facinfo !: "type" of
-             "randFun" -> Right (facinfo !: "id")
-             "uniformCategorical" ->
-               let n = facinfo !: "n" in
-               Left $ expFamFactor (categoricalValueExpFam n) [] (replicate (n - 1) 0.0, replicate (n-1) [])
-             "normal" ->
-               let mean = facinfo !: "mean"
-                   stdev = facinfo !: "stdev"
-              in Left $ expFamFactor gaussianValueExpFam [] ([mean / stdev^2, -1 / (2 * stdev^2)], [[]])
-             "constant" -> Left $ constFactor (decodeExpFam (facinfo !: "expFam")) $ case facinfo !: "expFam" !: "type" of
-               "gaussian" -> DoubleValue (facinfo !: "value")
-               "bernoulli" -> BoolValue (facinfo !: "value")
-               "categorical" -> CategoricalValue (facinfo !: "value")
[…]

### 2015-03-02 src/Quipp/ExpFam.hs

@@ -1,6 +1,6 @@
-{-# LANGUAGE RankNTypes, TypeFamilies, NoMonomorphismRestriction, ScopedTypeVariables #-}
+{-# LANGUAGE RankNTypes, TypeFamilies, NoMonomorphismRestriction, ScopedTypeVariables, ViewPatterns #-}
 
-module Quipp.ExpFam (ExpFam(ExpFam, expFamName, expFamD, expFamSufStat, expFamG, expFamSample, expFamRandomNatParam),
+module Quipp.ExpFam (ExpFam(ExpFam, expFamName, expFamD, expFamSufStat, expFamG, expFamSample, expFamDefaultNatParam, expFamRandomNatParam),
                      expFamFeaturesD, expFamSufStatToFeatures, expFamFeaturesToSufStat,
                      expFamSufStatToLikelihood, expFamCrossEntropy,
                      getNatParam,
@@ -10,7 +10,8 @@ module Quipp.ExpFam (ExpFam(ExpFam, expFamName, expFamD, expFamSufStat, expFamG,
                      likelihoodLogProbability,
                      timesLikelihood, productLikelihoods,
                      sampleLikelihood, expSufStat, covarianceSufStat,
-                     gaussianExpFam, categoricalExpFam) where
+                     gaussianExpFam, categoricalExpFam,
+                     expFamEntropy) where
 
 import Debug.Trace
 import Control.Monad (liftM, zipWithM)
@@ -33,6 +34,7 @@ data ExpFam v = ExpFam {
   expFamG :: forall s. RealFloat s => [s] -> s,
   expFamSufStatToLikelihood :: [Double] -> Likelihood v,
   expFamSample :: [Double] -> RVar v,
+  expFamDefaultNatParam :: [Double],
   expFamRandomNatParam :: RVar [Double],
   expFamFeaturesMask :: [Bool]
 }
@@ -66,6 +68,7 @@ promoteExpFam (f, finv) ef = ExpFam {
   expFamSufStatToLikelihood = fmap f . expFamSufStatToLikelihood ef,
   expFamSample = liftM f . expFamSample ef,
   expFamRandomNatParam = expFamRandomNatParam ef,
+  expFamDefaultNatParam = expFamDefaultNatParam ef,
   expFamFeaturesMask = expFamFeaturesMask ef
 }
 
@@ -150,6 +153,15 @@ sampleVariances xs = (var snd3, var thd3)
[…]

### 2015-03-02 src/Quipp/Factor.hs

@@ -76,7 +76,11 @@ getFeatureMessage (expFamName -> "gaussian") f [x] = [d - d2 * x, d2 / 2]
   where f' = f . (:[])
         d = diff f' x
         d2 = diff (diff f') x
-getFeatureMessage ef f x | "categorical" `isPrefixOf` expFamName ef = grad f x
+-- getFeatureMessage ef f x | "categorical" `isPrefixOf` expFamName ef = grad f x
+getFeatureMessage ef f x | "categorical" `isPrefixOf` expFamName ef =
+  let n = length x + 1
+      values = [f [if j == i then 1.0 else 0.0 | j <- [1 .. n - 1]] | i <- [0 .. n - 1]]
+  in map (\x -> x - head values) (tail values)
 getFeatureMessage ef f x = error $ "Bad expfam and features: " ++ expFamName ef ++ ", " ++ show x
 
 -- TODO: something is fishy here.  Things get flipped when they shouldn't.
@@ -92,8 +96,7 @@ expFamFactor ef argExpFams eta@(etaBase, etaWeights) =
   where expSufStatAndFeatures (likelihood:argsLikelihoods) = (expSufStat ef likelihood, concat [expFamSufStatToFeatures aef (expSufStat aef l) | (l, aef) <- zip argsLikelihoods argExpFams])
         fnp 0 (_, feats) = getNatParam ef eta feats
         fnp n (ss, feats) =
-          let gradProbNp = grad (\np -> dotProduct np (map fromDouble ss) - expFamG ef np) $ getNatParam ef eta feats
-              minFeatureIndex = sum $ map expFamFeaturesD $ take (n-1) argExpFams
+          let minFeatureIndex = sum $ map expFamFeaturesD $ take (n-1) argExpFams
               thisArgDim = expFamFeaturesD (argExpFams !! (n-1))
               probFromFeats :: RealFloat a => [a] -> a
               probFromFeats feats' =

### 2015-03-02 src/Quipp/Util.hs

@@ -125,8 +125,17 @@ linSolve :: Matrix Double -> [Double] -> [Double]
 linSolve mat d =
   matMulByVector (Mat.toLists $ pinv $ Mat.fromLists mat) d
 
+diagEntries :: Matrix a -> [a]
+diagEntries m
+  | length m == 0 || length m == length (head m) =
+    zipWith (!!) m [0..]
+  | otherwise = error $ "Cannot get diagonal entries of non-square matrix " ++ show (length m, length (head m))
+
 fromDouble :: Fractional a => Double -> a
 fromDouble = fromRational . toRational
+
+toDouble :: Real a => a -> Double
+toDouble = fromRational . toRational
 
 {- f(x) = ax^2 + bx + c
  - f'(x) = 2ax + b
@@ -152,3 +161,4 @@ variance xs = sum [(x-m)^2 | x <- xs] / fromIntegral (length xs) where m = mean 
 covariance xys = sum [(x - ux) * (y - uy) | (x, y) <- xys] / fromIntegral (length xys)
   where ux = mean (map fst xys)
         uy = mean (map snd xys)
+

### 2015-03-02 src/python/callhaskell.py

@@ -66,54 +66,33 @@ def varvals_to_state(varvals):
   for k in varvals:
     s.append([k, {u'Left': varvals[k]}])
 
-
-def temp_file_name(prefix=''):
-    return '/tmp/' + prefix + '_' + str(uuid.uuid1())
-
-def run_hs(*args):
-    subprocess.call(['../../dist/build/quipp_main/quipp_main'] + list(args))
-
-def dump_template(templ):
-    templ_file = temp_file_name('template')
-    json.dump(templ, open(templ_file, 'w'))
-    return templ_file
-
-def dump_state(state):
-    state_file = temp_file_name('state')
-    json.dump(state, open(state_file, 'w'))
-    return state_file
-
-def dump_params(params):
-    params_file = temp_file_name('params')
-    json.dump(params, open(params_file, 'w'))
-    return params_file
-
-def load_contents(fname):
-  return json.load(open(fname))
+def run_hs(command, *args):
+    popen = subprocess.Popen(['../../dist/build/quipp_main/quipp_main', command], stdin=subprocess.PIPE, stdout=subprocess.PIPE)
+    (output, _) = popen.communicate(json.dumps(list(args)))
+    return json.loads(output)
 
 def hs_rand_template_params(templ):
-    params_file = temp_file_name('params')
-    run_hs('randTemplateParams', dump_template(templ), params_file)
-    return load_contents(params_file)
+  return run_hs('randTemplateParams', templ)
 
 def hs_sample_bayes_net(templ, params):
-    state_file = temp_file_name('state')
-    run_hs('sampleBayesNet', dump_template(templ), dump_params(params), state_file)
-    return load_contents(state_file)
+  return run_hs('sampleBayesNet', templ, params)
 
 def hs_init_em(templ):
-    state_file = temp_file_name('state')
-    params_file = temp_file_name('params')
-    run_hs('initEM', dump_template(templ), state_file, params_file)
-    return (load_contents(state_file), load_contents(params_file))
+  return run_hs('initEM', templ)
 
-def hs_step_em(templ, state, params):
-    state_file = dump_state(state)
-    params_file = dump_params(params)
-    run_hs('stepEM', dump_template(templ), state_file, params_file)
-    return (load_contents(state_file), load_contents(params_file))
+def hs_infer_state(templ, state, params, iters=10):
+  return run_hs('inferState', templ, state, params, iters)
 
[…]

### 2015-03-02 src/python/examples/hmm.py

@@ -2,7 +2,7 @@ from quipp import *
 
 
 def run():
-  num_states = 3
+  num_states = 4
   num_observations = 3
 
   StateType = Categorical(num_states)

### 2015-03-02 src/python/graphbuilder.py

@@ -312,18 +312,24 @@ def conditioned_network(state, typ, sampler, frozen_samps):
     state.unify_values(get_type(s), s, unfrozen)
   return [latent for (latent, _) in samples]
 
-def infer_parameters_from_samples(graph_state, samples, frozen_samples):
+def condition_on_frozen_samples(graph_state, samples, frozen_samples):
   for s,f in zip(samples, frozen_samples):
     typ = get_type(s[1])
     graph_state.unify_values(typ, s[1], typ.unfreeze(graph_state, f))
-  templ = graph_state.to_JSON()
+  return graph_state.to_JSON()
+
+def infer_states_and_parameters(templ):
   (state, params) = hs_init_em(templ)
-  yield (state, params)
+  state = hs_infer_state(templ, state, params)
+  score = hs_score(templ, state, params)
+  yield (state, params, score)
   i = 0
   while True:
     print 'iter', i
-    (state, params) = hs_step_em(templ, state, params)
-    yield (state, params)
+    params = hs_infer_params(templ, state, params)
+    state = hs_infer_state(templ, state, params)
+    score = hs_score(templ, state, params)
+    yield (state, params, score)
     i += 1
 
 def translate_params_for_fn(params):
@@ -386,17 +392,20 @@ def run_clustering_example(run):
     frozen_samples = [freeze_value(samp, varvals) for samp in samples]
     true_latents = [x[0] for x in frozen_samples]
     print true_latents
-    state_params_list = infer_parameters_from_samples(current_graph_state, samples, [x[1] for x in frozen_samples])
+    templ = condition_on_frozen_samples(current_graph_state, samples, [x[1] for x in frozen_samples])
+    print 'best score', params_score(templ, rand_params)
+    state_params_list = infer_states_and_parameters(templ)
     rand_cs = params_to_cluster_centers(rand_params)
     iter_accs = []
     j = 0
-    for (state, params) in state_params_list:
+    for (state, params, score) in state_params_list:
+      print 'score', score
       cs = params_to_cluster_centers(params)
-      if j > 1:
-        varvals = state_to_varvals(state)
-        state_latents = [freeze_value(samp[0], varvals) for samp in samples]
-        acc = cluster_assignment_accuracy(true_latents, state_latents)
-        iter_accs.append(acc)
+      # if j > 1:
+      #   varvals = state_to_varvals(state)
[…]
